Determining the existance and direction of causality in the face of measurement error

### 2016-05-05 docs/manuscript.Rmd
updated

---
title: "Determining the existance and direction of causality in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  <!-- word_document -->
  pdf_document:
    keep_tex: true
bibliography: manuscript.bib
csl: nature.csl
---

@@

$$
\begin{aligned}
x & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_m + \beta_m x + \epsilon_m \\
y & = \alpha_x + \beta_x x + \epsilon_x
\end{aligned}
$$

where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, and $x_O$ is the measured quantity of $x$, where measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and imprecision from $\epsilon_m$, our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$. Under the CIT, this is expected to be $cov(g, y - \hat{y}) = 0$ when $x$ causes $y$, where $\hat{y} = \hat{a}_{x_O} + \hat{b}_{x_O} x_O + \epsilon_{x_O}$ is the predicted value of $y$ using the measured value of $x_O$.

We can split $cov(g, y - \hat{y})$ into two parts, $cov(g, y)$ and $cov(g, \hat{y})$.

@@

$$
\begin{aligned}
cov(g, y) & = cov(g, \beta_x x) \\
          & = cov(g, \beta_x \beta_g g) \\
          & = \beta_x\beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, \hat{y}) & = cov(g, \hat{\beta}_{x_O} x_O) \\
                & = cov(g, \hat{\beta}_{x_O} \beta_m x) \\
                & = cov(g, \hat{\beta}_{x_O} \beta_m \beta_g g) \\
                & = \hat{\beta}_{x_O} \beta_m \beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
\hat{\beta}_{x_O} & = \frac{cov(y, x_O)} {var(x_O)} \\
                  & = \frac{\beta_m cov(y, x)} {var(x_O)} \\
                  & = \frac{\beta_m \beta_x var(x)} {var(x_O)} \\
                  & = \frac{\beta_m \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_x\beta_g var(g)
\end{aligned}
$$

>>>>

---
title: "Determining the existance and direction of causality in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: true
bibliography: manuscript.bib
csl: nature.csl
---

@@

$$
\begin{aligned}
x   & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_m + \beta_m x + \epsilon_m \\
y   & = \alpha_x + \beta_x x + \epsilon_x \\
y_O & = \alpha_n + \beta_n y + \epsilon_n
\end{aligned}
$$

where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and imprecision from $\epsilon_m$, and $y_O$ is the measured quantity of $y$, where measurement error is incurred from bias in $\alpha_n$ and $\beta_n$ and imprecision from $\epsilon_n$, our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$. Under the CIT, this is expected to be $cov(g, y_O - \hat{y_O}) = 0$ when $x$ causes $y$, where $\hat{y}_O = a_n + \beta_n(\hat{a}_{x_O} + \hat{\beta}_{x_O} x_O)$ is the predicted value of $y_O$ using the measured value of $x_O$.

We can split $cov(g, y_O - \hat{y}_O)$ into two parts, $cov(g, y_O)$ and $cov(g, \hat{y}_O)$.

@@

$$
\begin{aligned}
cov(g, y_O) & = cov(g, \beta_n y) \\
            & = cov(g, \beta_n \beta_x x) \\
            & = cov(g, \beta_n \beta_x \beta_g g) \\
            & = \beta_n \beta_x \beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, \hat{y}_O) & = cov(g, \alpha_n + \beta_n \alpha_x + \beta_n \hat{\beta}_{x_O} x_O) \\
                  & = cov(g, \beta_n \hat{\beta}_{x_O} \beta_m x) \\
                  & = cov(g, \beta_n \hat{\beta}_{x_O} \beta_m \beta_g g) \\
                  & = \hat{\beta}_{x_O} \beta_n \beta_m \beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
\hat{\beta}_{x_O} & = \frac{cov(y_O, x_O)} {var(x_O)} \\
                  & = \frac{cov(\beta_n y, \beta_m x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                  & = \frac{\beta_m \beta_n cov(y, x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                  & = \frac{\beta_m \beta_n \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, \hat{y}_O) & = \frac{\beta_n \beta_x \beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                  & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_x\beta_g var(g)
\end{aligned}
$$
